Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-23 manuscript.tex
mod conclusions
update runtimes by Leonardo

Minimisation of the nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (MacBook Air, dual core 1.6 GHz i5 processor, 16 GB RAM). Compared to Leiden approach, nSBM requires at best $\sim8\times$ times more, and $\sim180\times$  at worst (possibly due to limited memory for large networks). A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:08 & 00:55 & 01:23 \\ 
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:24 & 01:34 & 03:29\\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:16 & 00:30 & 02:10 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:24 & 01:10 & 06:11 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:34 & 01:31 & 09:06 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:34 & 02:53 & 06:55\\ 
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 06:32 & 34:12 & 3:17:08 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. Times are expressed in mm:ss. }
\label{Table_runtime}
\end{table}

@@

Identification of cells sharing similar properties in single cell experiments is of paramount importance. A large number of approaches have been described, although the standardisation of analysis pipelines converged to methods that are based on modularity optimisation. We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques. The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution. In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble. \textcolor{red}{Our approach introduces the evaluation of cluster consistency, which can be used to isolate cells with heterogeneous identity}.

>>>>

Minimisation of the nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimize multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (\textcolor{red}{dual core 1.6 GHz i5 processor, 16 GB RAM}). Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst. A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

\begin{table}[h!]
\centering
 \begin{tabular}{|| l c c c c c ||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
 \textbf{Dataset} & \textbf{Cells} & \textbf{Edges} & \textbf{Leiden} & \textbf{PPBM} & \textbf{nSBM} \\ [0.5ex] 
 \hline\hline
 sc-mixology \cite{Tian_2019} & 860 & 9186 & 00:06 & 00:13 & 00:36 \\ 
 \hline
 Quartz-seq2 \cite{mereu_2020} & 1266 & 14603 & 00:10 & 00:19 & 00:45 \\
 \hline
 MARS-seq \cite{mereu_2020} & 1401 & 21756 & 00:20 & 00:34 & 02:14 \\
 \hline
 iCELL8 \cite{mereu_2020} & 1830 & 30636 & 00:23 & 00:40 & 03:02 \\
 \hline
 Chromium 10x \cite{mereu_2020} & 1523 & 21447 & 00:14 & 00:26 & 01:07 \\ 
 \hline
 Islet \textcolor{red}{??} & 2108 & 32023 & 00:20 & 00:42 & 03:13 \\
 \hline
 Hematopoiesis \cite{paul_2015} & 2730 & 15444 & 00:37 & 01:27 & 05:52\\ 
 \hline
 Zeisel Cortex \textcolor{red}{??} & 3005 & 54460 & 00:59 & 00:53 & 07:32 \\
 \hline
 \textcolor{red}{Pancreas velocity} \textcolor{red}{??} & 3696 & 74670 & 01:15 & 01:29 & 10:56 \\
 \hline
 Baron Pancreas \textcolor{red}{??} & 8569 & 294480 & 03:51 & 07:35 & 1:33:40 \\
 \hline
 Airzani Liver \textcolor{red}{??} & 10368 & 354440 & 04:13 & 09:47 & 1:42:23 \\
 \hline
 Planaria \cite{plass_2018} & 21612 & 173667 & 05:41 & 13:52 & 1:20:40 \\
 \hline
\end{tabular}
\caption{Time required to run different partitioning strategies implemented in \texttt{schist} on various datasets. All approaches fit 100 models. Number of nodes and edges refer to the structure of the \emph{k}NN graph as built by \texttt{scanpy}. Times are expressed in mm:ss. }
\label{Table_runtime}
\end{table}

@@

Identification of cells sharing similar properties in single cell experiments is of paramount importance. A large number of approaches have been described, although the standardisation of analysis pipelines converged to methods that are based on modularity optimisation. We tackled the biological problem using a different approach, nSBM, which has several advantages over existing techniques. The most important advantage is the hierarchical definition of cell groups which eliminates the choice of an arbitrary threshold on clustering resolution. In addition, we showed that the hierarchy itself could have a biological interpretation, implying that the hierarchical model is a valid representation of the cell ensemble.
